---
title: "Senior Project"
author: "Jorge Jeronimo Fueyo Gonzalez"
date: "3/10/2022"
output:
  pdf_document: default
  html_document: default
---

# S&DS Senior Project (S&DS 492)
## Jorge Jeronimo Fueyo Gonzalez
## Advisor: Professor Meyers
### Spring 2022

## Senior Project











Three major topics:

1. Factors that Contribute to Violence in Mexico

2. Effects of Violence on the Economic Development of Mexico

3. Economic Phenomenon of Collusion: Why do Cartels Collude and Cheat on Each Other?


```{r install}
# install packages
#install.packages("readxl")
#install.packages("tidyverse")
```

```{r load}
# load packages
library("readxl")
library("tidyverse")
library("dplyr")
```


```{r violence_data_load}
# load data

df1 <- read.csv("/Users/jeronimofueyo/Documents/Jeronimo Fueyo/RStudio/Yale-SDS-492-Senior-Project/21-04-22 10_51_42_theglobaleconomy.csv")
df2 <- read.csv("/Users/jeronimofueyo/Documents/Jeronimo Fueyo/RStudio/Yale-SDS-492-Senior-Project/21-04-22 11_13_01_theglobaleconomy.csv")

```

```{r}
View(df1)
View(df2)
```


```{r}
df3 <- df1 %>%
  left_join(df2, by="Year") %>%
  rename(year = Year, gini = Gini.income.inequality.index, poverty_percent = Poverty.percent.of.population,
         education_spending = Public.spending.on.education.percent.of.public.spending, literacy_rate = Literacy.rate,
         secondary_school_enrolment = Secondary.school.enrollment.percent.of.all.eligible.children,
         tertiary_school_enrolment = Tertiary.school.enrollment.percent.of.all.eligible.children,
         globalization = Globalization.index..0.100., econ_globalization = Economic.globalization.index..0.100.,
         GDP_prcnt_wrld = Percent.of.world.GDP, FDI_prcnt_wrld = Percent.of.world.Foreign.Direct.Investment,
         savings_bUSD = Savings.billion.USD, intl_tourism_rev = International.tourism.revenue,
         kidnappings_per_100K = Kidnappings.per.100000.people, robberies_per_100K = Robberies.per.100000.people,
         prisoners_per_100K = Number.of.prisoners.per.100000.people, thefts_per_100K = Thefts.per.100000.people,
         homicides_per_100K = Homicides.per.100000.people,
         military_spending_bUSD = Military.spending.in.billion.U.S..dollars,
         cost_of_living = Cost.of.living.index.world.average...100,
         education_price = Education.service.price.index.world.average...100,
         GDP_change_rate = Economic.growth..the.rate.of.change.of.real.GDP,
         GDP_bUSD = Gross.Domestic.Product.billions.of.U.S..dollars, GDP_PC = GDP.per.capita.current.U.S..dollars,
         GDP_PC_PPP = GDP.per.capita.Purchasing.Power.Parity, capital_investment = Capital.investment.billion.USD,
         inlfation = Inflation..percent.change.in.the.Consumer.Price.Index, labor_force = Labor.force.million.people,
         unemployment_rate = Unemployment.rate, trade = Terms.of.trade.base.year...2000,
         gvt_spending = Government.spending.billion.USD, gvt_debt = Government.debt.as.percent.of.GDP,
         rule_of_law = Rule.of.law.index...2.5.weak..2.5.strong.,
         corruption_control = Control.of.corruption...2.5.weak..2.5.strong.,
         corruption_perception = Corruption.Perceptions.Index.100...no.corruption,
         competitiveness = Competitiveness...World.Economic.Forum.index..0.100.,
         cost_start_business = Cost.of.starting.a.business...of.income.per.capita,
         shadow_economy_pct_GDP = Shadow.economy.percent.of.GDP, population_M = Population.size.in.millions,
         migrant_pop_pct = Migrant.population.percent.of.total.population,
         health_spending_PC = Health.spending.per.capita) %>%
  select(year, gini, education_spending, literacy_rate, secondary_school_enrolment, tertiary_school_enrolment,
         globalization, econ_globalization, GDP_prcnt_wrld, FDI_prcnt_wrld, savings_bUSD, intl_tourism_rev,
         kidnappings_per_100K, robberies_per_100K, prisoners_per_100K, thefts_per_100K, homicides_per_100K,
         military_spending_bUSD, GDP_change_rate, GDP_bUSD, GDP_PC, GDP_PC_PPP, capital_investment, inlfation,
         labor_force, unemployment_rate, trade, gvt_spending, gvt_debt, rule_of_law, corruption_control,
         corruption_perception, cost_start_business, shadow_economy_pct_GDP, population_M, health_spending_PC) %>%
  mutate(FDI_prcnt_wrld_lag = lag(FDI_prcnt_wrld, n=1), intl_tourism_rev_lag = lag(intl_tourism_rev, n=1),
         military_spending_bUSD_lag = lag(military_spending_bUSD, n=1), gvt_spending_lag = lag(gvt_spending, n=1))
  
  
```

```{r}
View(df3)
```

```{r}
head(df3)
```


```{r}
fit1 <- lm(homicides_per_100K ~ unemployment_rate, df3)
summary(fit1)
```


```{r}
fit2 <- lm(homicides_per_100K ~ literacy_rate, df3)
summary(fit2)
```

```{r}
fit3 <- lm(homicides_per_100K ~ unemployment_rate + education_spending, df3)
summary(fit3)
```


```{r}
fit4 <- lm(intl_tourism_rev_lag ~ homicides_per_100K, df3)
summary(fit4)
```

```{r}
fit5 <- lm((prisoners_per_100K/homicides_per_100K) ~ corruption_perception, df3)
summary(fit5)
```

```{r}
fit6 <- lm(GDP_PC ~ homicides_per_100K + unemployment_rate + FDI_prcnt_wrld + trade, df3)
summary(fit6)
```


```{r}
ggplot(df3, aes(x=homicides_per_100K, y=military_spending_bUSD, color=year)) +
  geom_point() +
  geom_smooth(method="lm")
```



#### Factors that Contribute to Violence in Mexico

```{r violence_data_load}
# load violence data

df_homicides_0 <- read.csv("INEGI_exporta_10_2_2022_13_0_2.csv")

```



1. How does level of education affects level of violence in a specific region (state)?

  Relationship between level of education and:
  
  a) homicides
  b) feminicides
  c) crime rate
  d) kidnappings
  
  - Homicides data is readily available
  - Data on crime rate and kidnappings might be harder to find

2. What's the relationship between the political party in power in a region (state) with the level of violence in that region?

  - Data set would have to be made by me based on information online
  - Try to find time series data set

3. What's the relationship between the drug cartel in power in a region (state) with level of violence in that region?

  - Data set would have to be made by me based on information online
  - Try to find time series data set

4. Is there a pattern between both, the political party and drug cartel in power in a region (state) and their trends on violence?


#### Effects of Violence on the Economic Development of Mexico


```{r economic_data_load}
# load economic data

df_GDP_PC_0 <- read.csv("/Users/jeronimofueyo/Documents/Jeronimo Fueyo/GitHub/Yale-SDS-492-Senior-Project/PCAGDPMXA646NWDB.csv")

df_GDP_0 <- read.csv("/Users/jeronimofueyo/Documents/Jeronimo Fueyo/GitHub/Yale-SDS-492-Senior-Project/MKTGDPMXA646NWDB.csv")

df_FDI_0 <- read.csv("/Users/jeronimofueyo/Documents/Jeronimo Fueyo/GitHub/Yale-SDS-492-Senior-Project/BPFADI03MXQ637N.csv")

```


```{r}
df_GDP_PC <- df_GDP_PC_0 %>%
  rename(date = DATE, GDP_PC = PCAGDPMXA646NWDB) %>%
  mutate(date = as.Date(date), year = as.numeric(format(date, format="%Y")))

df_GDP <- df_GDP_0 %>%
  rename(date = DATE, GDP = MKTGDPMXA646NWDB) %>%
  mutate(date = as.Date(date), year = as.numeric(format(date, format="%Y")))

df_FDI <- df_FDI_0 %>%
  rename(date = DATE, FDI = BPFADI03MXQ637N) %>%
  mutate(date = as.Date(date), year = as.numeric(format(date, format="%Y")))

df_FDI_per_year <- df_FDI %>%
  group_by(year) %>%
  summarize(FDI = mean(FDI))
```


```{r economic_data_head}
# head economic data

head(df_GDP_PC)
head(df_GDP)
head(df_FDI)
head(df_FDI_per_year)
```

```{r economic_data_view}
# view economic data

View(df_GDP_PC)
View(df_GDP)
View(df_FDI)
View(df_FDI_per_year)
```


```{r}
df_homicides <- df_homicides_0[6:36,] %>%
  select(!X.35) %>%
  rename(year = Defunciones.por.homicidios, total_homicides = X,
         Aguascalientes_homicides = X.1, BajaCalifornia_homicides = X.2,
         BajaCaliforniaSur_homicides = X.3, Campeche = X.4,
         Coahuila = X.5, Colima = X.6,
         Chiapas = X.7, Chihuahua = X.8,
         CDMX = X.9, Durango = X.10,
         Guanajuato = X.11, Guerrero = X.12,
         Hidalgo = X.13, Jalisco = X.14,
         Mexico = X.15, Michoacan = X.16,
         Morelos = X.17, Nayarit = X.18,
         NuevoLeon = X.19, Oaxaca = X.20,
         Puebla = X.21, Queretaro = X.22,
         QuintanaRoo = X.23, SanLuisPotosi = X.24,
         Sinaloa = X.25, Sonora = X.26,
         Tabasco = X.27, Tamaulipas = X.28,
         Tlaxcala = X.29, Veracruz = X.30,
         Yucatan = X.31, Zacatecas = X.32,
         Extranjero = X.33, NotIdentified = X.34) %>%
  mutate(year = as.numeric(year),
         total_homicides = as.numeric(gsub(",", "", total_homicides)),
         Aguascalientes_homicides = as.numeric(gsub(",", "", Aguascalientes_homicides)),
         BajaCalifornia_homicides = as.numeric(gsub(",", "", BajaCalifornia_homicides)),
         BajaCaliforniaSur_homicides = as.numeric(gsub(",", "", BajaCaliforniaSur_homicides)),
         Campeche = as.numeric(gsub(",", "", Campeche)),
         Coahuila = as.numeric(gsub(",", "", Coahuila)),
         Colima = as.numeric(gsub(",", "", Colima)),
         Chiapas = as.numeric(gsub(",", "", Chiapas)),
         Chihuahua = as.numeric(gsub(",", "", Chihuahua)),
         CDMX = as.numeric(gsub(",", "", CDMX)),
         Durango = as.numeric(gsub(",", "", Durango)),
         Guanajuato = as.numeric(gsub(",", "", Guanajuato)),
         Guerrero = as.numeric(gsub(",", "", Guerrero)),
         Hidalgo = as.numeric(gsub(",", "", Hidalgo)),
         Jalisco = as.numeric(gsub(",", "", Jalisco)),
         Mexico = as.numeric(gsub(",", "", Mexico)),
         Michoacan = as.numeric(gsub(",", "", Michoacan)),
         Morelos = as.numeric(gsub(",", "", Morelos)),
         Nayarit = as.numeric(gsub(",", "", Nayarit)),
         NuevoLeon = as.numeric(gsub(",", "", NuevoLeon)),
         Oaxaca = as.numeric(gsub(",", "", Oaxaca)),
         Puebla = as.numeric(gsub(",", "", Puebla)),
         Queretaro = as.numeric(gsub(",", "", Queretaro)),
         QuintanaRoo = as.numeric(gsub(",", "", QuintanaRoo)),
         SanLuisPotosi = as.numeric(gsub(",", "", SanLuisPotosi)),
         Sinaloa = as.numeric(gsub(",", "", Sinaloa)),
         Sonora = as.numeric(gsub(",", "", Sonora)),
         Tabasco = as.numeric(gsub(",", "", Tabasco)),
         Tamaulipas = as.numeric(gsub(",", "", Tamaulipas)),
         Tlaxcala = as.numeric(gsub(",", "", Tlaxcala)),
         Veracruz = as.numeric(gsub(",", "", Veracruz)),
         Yucatan = as.numeric(gsub(",", "", Yucatan)),
         Zacatecas = as.numeric(gsub(",", "", Zacatecas)),
         Extranjero = as.numeric(gsub(",", "", Extranjero)),
         NotIdentified = as.numeric(gsub(",", "", NotIdentified))) %>%
  replace(is.na(.), 0)
```


```{r homicides_data_head}
# head homicides data

head(df_homicides)
```

```{r homicides_data_view}
# view homicides data

View(df_homicides)
```

```{r}
df_violence_econ <- df_homicides %>%
  left_join(df_GDP, by="year") %>%
  left_join(df_GDP_PC, by="year") %>%
  left_join(df_FDI_per_year, by="year") %>%
  select(year, total_homicides, GDP, GDP_PC, FDI) %>%
  mutate(years_since_1989 = year - 1989)

head(df_violence_econ)
```



``` {r}
ggplot(df_violence_econ, aes(x=total_homicides, y=GDP, color=year)) +
  geom_point() +
  geom_smooth(method="lm")
```

``` {r}
ggplot(df_violence_econ, aes(x=total_homicides, y=GDP_PC, color=year)) +
  geom_point() +
  geom_smooth(method="lm")
```

``` {r}
ggplot(df_violence_econ, aes(x=total_homicides, y=FDI, color=year)) +
  geom_point() +
  geom_smooth(method="lm")
```


``` {r}
ggplot(df_violence_econ, aes(x=year, y=GDP)) +
  geom_point() +
  geom_line() +
  geom_smooth(method="lm")
```

``` {r}
ggplot(df_violence_econ, aes(x=year, y=total_homicides)) +
  geom_point() +
  geom_line()
```

``` {r}
fit_1 <- lm(total_homicides ~ years_since_1989 + GDP, df_violence_econ)
summary(fit_1)
```

``` {r}
fit_2 <- lm(total_homicides ~ years_since_1989 + GDP_PC, df_violence_econ)
summary(fit_2)
```


``` {r}
fit_3 <- lm(total_homicides ~ years_since_1989 + FDI, df_violence_econ)
summary(fit_3)
```

``` {r}
fit_3 <- lm(total_homicides ~ years_since_1989 + GDP_PC + FDI, df_violence_econ)
summary(fit_3)
```


```{r}
fit_4 <- lm(FDI ~ years_since_1989 + total_homicides, df_violence_econ)
summary(fit_4)
```


```{r}
df_violence_econ_not_90s <- df_violence_econ %>%
  mutate(years_since_1999 = year - 1999) %>%
  filter(year >= 2000)
head(df_violence_econ_not_90s)
```

``` {r}
ggplot(df_violence_econ_not_90s, aes(x=total_homicides, y=GDP_PC, color=year)) +
  geom_point() +
  geom_smooth(method="lm")
```


```{r}
fit_5 <- lm(total_homicides ~ years_since_1999 + GDP_PC, df_violence_econ_not_90s)
summary(fit_5)
```


```{r}
fit_6 <- lm(total_homicides ~ years_since_1989 + GDP_PC + FDI + GDP_PC*FDI, df_violence_econ)
summary(fit_6)
```


1. What's the effect of violence (homicides/crime rates) on economic development in the country as a whole?
  
  Effect of violence (homicides/crime rates) on:
  
  a) GDP/GDP per capita?
    i) Already have the data
  b) foreign investment
    i) Already have the data

2. What's the effect of violence (homicides/crime rates) on employment and business proliferation?

  Effect of violence on:
  
  a) level of unemployment per region
    i) not sure if I can get data on unemployment per region each year
  b) number of business created per region
    i) might be hard to get data on this


#### Economic Phenomenon of Collusion: Why do Cartels Collude and Cheat on Each Other?

In Microeconomics, we see that firms have an incentive to collude and sometimes cheat on each other. This seldom happens in regulated markets. The drug market through which these criminal groups operate is illegal, which makes it unregulated. Drug cartels are incentivized to collude and form aliances, but sometimes also cheat on each other and create war, which results in more violence. In this part of the project, I aim to discuss this phenomenon and how it has affected levels of violence in the country.

- I see it hard to find a way to use data to support this discussion so I don't think I'll be using data here


